Partial rank data with the \pkg{hyper2} package: likelihood functions for generalized Bradley-Terry models

### 2019-05-03 inst/collusion_chess_draws.Rnw
more work

\newcommand{\bolda}{\mathbf a}
\newcommand{\boldp}{\mathbf p}
\newcommand{\boldV}{\mathbf V}
\newcommand{\boldalpha}{\mbox{\boldmath$\alpha$}}
\newcommand{\boldzero}{\mathbf 0}

\newcommand{\mcf}{\mathcal F}
\newcommand{\mcm}{\mathcal M}
\newcommand{\mcs}{\mathcal S}
\newcommand{\mcw}{\mathcal W}

\newcommand{\pG}{p_{\mbox{\frakfamily g}}}

@@

where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
the number of times $i$ loses to $j$.  Incorporation of draws is
straightforward: we consider a chess match to be a three-way trial
between the two players and an entity that wins if the match is drawn.
We thus seek a likelihood function for non-negative
$p_1,\ldots,p_n,p_D$ $p_D+\sum p_i=1$ and this would be

@@

\begin{equation}
\prod_{i<j}
\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}
\left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}
\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\end{equation}

@@

which might be preferable on grounds of clarity.  We now apply this
likelihood model to a real dataset.

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays white, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  13 &   2  \\
  Kr &  5 &  -  &   7  \\
  Ks & 11 &   9 &  -   \\
\end{tabular}
\caption{Results of 365 chess matches \label{kka} between three
  Grandmasters.  Thus, for example, Anand played white against Karpov
  a total of $18+20+13=51$ times and won 18, drew 20, lost 13}
\end{table}

The likelihood function is

>>>>

@@

where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
the number of times $i$ loses to $j$.  Incorporation of draws is
straightforward: we consider a chess match to be a three-way trial
between the two players and an entity that wins if the match is drawn.
We thus seek a likelihood function for non-negative
$p_1,\ldots,p_n,p_D$ with $p_D+\sum p_i=1$ and this would be

@@

\begin{align}\nonumber
\prod_{i<j}&{}
\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}\\
&{} \left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_kka}
\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}\\
&{} \left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\nonumber
\end{align}

@@

which might be preferable on grounds of clarity.  We now apply this
likelihood model to a real dataset, shown in table~\ref{kka}.  The
likelihood function is easily maximized,
figure~\ref{maxp_karpov_kasparov_anand}.

\begin{figure}[htbp]
\begin{center}
<<rowing_maxp,echo=TRUE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_kka}}
  \end{center}
\end{figure}

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  13 &   2  \\
  Kr &  5 &  -  &   7  \\
  Ks & 11 &   9 &  -   \\
\end{tabular}
\caption{Results of 365 chess matches \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  Thus, for
  example, Anand played white against Karpov a total of $18+20+13=51$
  times and won 18, drew 20, lost 13}
\end{table}

Figure~\ref{kka} suggests a number of interesting hypotheses.
Firstly, we observe that $p_D=0$ cannot possibly be the case: the
likelihood would be identically zero, owing to the nonzero term in the
numerator of the last term of Equation~\ref{likelihood_kka}.  However,
it is conceivable that $p_W=0$ and this hypothesis can readily be
tested using the concept of profile likelihood.
